A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-03-28 paper_comparison/paper_comparison/paper_comparison.Rmd
update intro

More typically, residuals will be plotted, as a scatter plot against the predicted values and each of the explanatory variables to scrutinize their relationships. If there are any visually discoverable patterns, the model is potentially misspecified. In general, one looks for noticeable departures from the model like non-linear dependency or heteroskedasticity. However, correctly judging a residual plot where no pattern exists can be a painstakingly difficult task for humans (?citation). It is especially common, particularly among new data analysts, to report patterns when an experienced one might quickly conclude that there are none (ET: not sure if this is true). It is also possible to conduct hypothesis tests for non-linear dependence [@ramsey_tests_1969], and use a Breusch-Pagan test [@breusch_simple_1979] for heteroskedasticity.

>>>>

More typically, residuals will be plotted, as a scatter plot against the predicted values and each of the explanatory variables to scrutinize their relationships. If there are any visually discoverable patterns, the model is potentially misspecified. In general, one looks for noticeable departures from the model like non-linear dependency or heteroskedasticity. However, correctly judging a residual plot where no pattern exists can be a painstakingly difficult task for humans. It is especially common, particularly among students, to misinterpret patterns that are "random noise" and "random deviation from a model" [@loy2021bringing]. It is also possible to conduct hypothesis tests for non-linear dependence [@ramsey_tests_1969], and use a Breusch-Pagan test [@breusch_simple_1979] for heteroskedasticity.

### 2023-03-28 paper_comparison/paper_comparison/paper_comparison.tex
update intro

More typically, residuals will be plotted, as a scatter plot against the
predicted values and each of the explanatory variables to scrutinize
their relationships. If there are any visually discoverable patterns,
the model is potentially misspecified. In general, one looks for
noticeable departures from the model like non-linear dependency or
heteroskedasticity. However, correctly judging a residual plot where no
pattern exists can be a painstakingly difficult task for humans
(?citation). It is especially common, particularly among new data
analysts, to report patterns when an experienced one might quickly
conclude that there are none (ET: not sure if this is true). It is also
possible to conduct hypothesis tests for non-linear dependence
\citep{ramsey_tests_1969}, and use a Breusch-Pagan test
\citep{breusch_simple_1979} for heteroskedasticity.

@@

The effectiveness of lineup protocol for regression analysis is
validated by \citet{majumder_validation_2013} under relatively simple
settings with up to two predictors. Their results suggest that visual
tests are capable of testing the significance of a single predictor with
a similar power as a t-test, though they express that in general it is
unnecessary to use visual inference if there exists a conventional test,
and they do not expect the visual test to perform equally well as the
conventional test. In their third experiment, where there is not a
conventional test, visual test outperforms the conventional test for a
large margin. This is encouraging, as it promotes the use of visual
inference in situations where there are no existing statistical testing
procedures. Visual inference have also been integrated into diagnostic
of hierarchical linear models by \citet{loy2013diagnostic},
\citet{loy2014hlmdiag} and \citet{loy2015you}. They use lineup protocols
to judge the assumption of linearity, normality and constant error
variance for both the level-1 and level-2 residuals. (expand?)

@@

A primary factor contributes to the non-linearity model is the shape of
the non-linearity pattern. According to Figure
\ref{fig:poly-power-uniform-j}, conventional tests have higher power in
testing shapes constructed from lower orders of Hermite polynomials. But
the chance of detecting the ``triple-U'' shape drops significantly
compared to other shapes. To understand why this is, one needs to return
to the way the RESET test is applied. It requires a parameter indicating
degree of fitted values to test for, and the recommendation is to
generically use four (\citet{ramsey_tests_1969}). However, the
``triple-U'' shape constructed from the Hermite polynomials use power up
to 18. If the RESET test had been applied using a higher power no less
than six, the power curve of ``triple-U'' shape will be only slight
lower than the power curve of ``M'' shape. The recommendation of the
polynomial power for the RESET test should be revised, perhaps. This
illustrates the sensitivity of conventional testing to the parameters,
and it also points to a limitation that one needs to know the data
generating process in order to set the parameters for the test.

@@

Although it is unusual to consider fitted value distribution or
predictor distribution in power analysis, it is not so surprised that
the power of conventional tests vary because of fitted value
distribution. For instance, the power of the RESET test which is
effectively a F test depends on the realizations of the additional
predictors included in the auxiliary regression equation
\citep{jamshidian2007study}. And those additional predictors are linear
transformation of random vectors following certain predictor
distributions. What is truly unexpected is the huge variability in power
of conventional tests compared to visual tests. The discrete uniform
distribution constantly produce the highest power while the normal
distribution and the lognormal distribution produce the lowest power for
non-linearity and heteroskedasticity pattern respectively. The
difference in power corresponding to different distributions peak at
around 90\% when the effect size is at a moderate level. This huge
difference is undesirable as is not uncommon for analysts to omit the
factor of predictor distribution in power calculation and sample size
calculation. It also suggests that, under the designed scenarios in this
paper, the visual test is a more robust test against fitted value
distributions than a conventional test.

>>>>

More typically, residuals will be plotted, as a scatter plot against the
predicted values and each of the explanatory variables to scrutinize
their relationships. If there are any visually discoverable patterns,
the model is potentially misspecified. In general, one looks for
noticeable departures from the model like non-linear dependency or
heteroskedasticity. However, correctly judging a residual plot where no
pattern exists can be a painstakingly difficult task for humans. It is
especially common, particularly among students, to misinterpret patterns
that are ``random noise'' and ``random deviation from a model''
\citep{loy2021bringing}. It is also possible to conduct hypothesis tests
for non-linear dependence \citep{ramsey_tests_1969}, and use a
Breusch-Pagan test \citep{breusch_simple_1979} for heteroskedasticity.

@@

The effectiveness of lineup protocol for regression analysis is
validated by \citet{majumder_validation_2013} under relatively simple
settings with up to two predictors. Their results suggest that visual
tests are capable of testing the significance of a single predictor with
a similar power as a t-test, though they express that in general it is
unnecessary to use visual inference if there exists a conventional test,
and they do not expect the visual test to perform equally well as the
conventional test. In their third experiment, where there is not a
conventional test, visual test outperforms the conventional test for a
large margin. This is encouraging, as it promotes the use of visual
inference in situations where there are no existing statistical testing
procedures. Visual inference have also been integrated into diagnostic
of hierarchical linear models by \citet{loy2013diagnostic},
\citet{loy2014hlmdiag} and \citet{loy2015you}. They use lineup protocols
to judge the assumption of linearity, normality and constant error
variance for both the level-1 and level-2 residuals.

@@

A primary factor contributes to the non-linearity model is the shape of
the non-linearity pattern. According to Figure
\ref{fig:poly-power-uniform-j}, conventional tests have higher power in
testing shapes constructed from lower orders of Hermite polynomials. But
the chance of detecting the ``triple-U'' shape drops significantly
compared to other shapes. To understand why this is, one needs to return
to the way the RESET test is applied. It requires a parameter indicating
degree of fitted values to test for, and the recommendation is to
generically use four \citep{ramsey_tests_1969}. However, the
``triple-U'' shape constructed from the Hermite polynomials use power up
to 18. If the RESET test had been applied using a higher power no less
than six, the power curve of ``triple-U'' shape will be only slight
lower than the power curve of ``M'' shape. The recommendation of the
polynomial power for the RESET test should be revised, perhaps. This
illustrates the sensitivity of conventional testing to the parameters,
and it also points to a limitation that one needs to know the data
generating process in order to set the parameters for the test.

@@

Although it is unusual to consider fitted value distribution or
predictor distribution in power analysis, it is not so surprised that
the power of conventional tests vary because of fitted value
distribution. For instance, the power of the RESET test which is
effectively a F test depends on the realizations of the additional
predictors included in the auxiliary regression equation
\citep{jamshidian2007study}. And those additional predictors are linear
transformation of random vectors following certain predictor
distributions. What is truly unexpected is the huge variability in power
of conventional tests compared to visual tests. The discrete uniform
distribution constantly produce the highest power while the normal
distribution and the lognormal distribution produce the lowest power for
non-linearity and heteroskedasticity pattern respectively. The
difference in power corresponding to different distributions peak at
around 90\% when the effect size is at a moderate level. This huge
difference is undesirable as is not uncommon for analysts to omit the
factor of predictor distribution in power calculation and sample size
calculation.
